---
title: "NUN - Introduction to Data Science Using R"
author: "Bilikisu Olatunji, Data Scientist, Abuja R User Group"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Brief introduction to R

## Importance of R

  - Open source
  - Robust visualization
  - Best and easy language for Statistical Analysis and Data Science
  - Vast Community support
  - Large ecosystems of packages
  
## Installing R and RStudio  


## Codes and Comments

```{r}
2+3
```

## Data Types 
  - Numeric
  - Integer
  - Logical
  - Character
  - Complex, Raw

```{r}
2+3

1
2.3
1L

# numeric
class(2.3)

# integer
class(1L)
# character
class("1")
class('34')
"23"+2

# convert a number to integer using as.integer()
as.integer(1)
class(as.integer(1))
class(1)
```


## Variables

```{r}
my_integer = 102L
# change the value
my_integer <- 107L
```


## Operators in R

  - Arithmetric
  - Assignment
  - Logical
  - Relational
  
```{r}
2+3

2*3

3/2

5^2
```

## Loops in R
  - for
  
```{r}
for (i in 1:5) {
  print(i)
}

```

  - while
  
```{r}
i <- 0
while (i <= 4) {
  i <- i + 1
  print(i)
}
```


## Data Structures in R

  1. Vectors
```{r}
c()

#Assign results to numeric_vec
numeric_vec <- c(2, 43, 4.1, 3.5, 10)

# Check the type and contents of numeric_vec
# We see that it is numeric (num)
str(numeric_vec)

# length
length(numeric_vec)
```

  2. Data frame
```{r}
# Let's create a simple data frame with the data.frame function
student_data <- data.frame(student_id = c(100234, 132454, 453123),
                      hw_score = c(82, 93, 87),
                      exam_score = c(92, 89, 98),
                      department = c("Statistics", "Physics", "Economics"))
# View the data
student_data
```
  
  3. List
```{r}
my_list <- list(char_vector = c('A', 'B', "Z"),
                numeric_vector = c(5.2, 13.4, 5, 2.01),
                student_data = data.frame(student_id = c(100234, 132454, 453123),
                      hw_score = c(82, 93, 87),
                      exam_score = c(92, 89, 98),
                      department = c("Statistics", "Physics", "Economics")),
                a_list = list(c(1L, 4L), c('A', 'D', 'E')))
# View contents
my_list
```
  

## Functions

### Base or built-in functions

```{r}
#c()

data_vector <- c(1, 3, 3, 9, 11.2, 14, 28.7, 30, 15, 5, 5, 5, 2.7, 21, 5.7, 9.1, 24.6)

sum(data_vector)
# 
mean(data_vector)

# 
median(data_vector)
# 
# min()
# 
# max()
```

### Custom functions
```{r}
# Define your custom functions
# function without an argument

myFunc <- function(){
  2+3
}

myFunc()

sayHello <- function(){
  print("Hello! This is my first function!")
}

sayHello()

# function with a single argument
multiplyByTwo <- function(x) {
  x*2
  #return(x * 2)
}

v1 <- c(1, 4, 7)

multiplyByTwo(x = v1)

multiplyByTwo(v1)
```

## Packages in R



### Installing packages

```{r}
# install.packages()

#install.packages("tidyverse")
```

### Accessing and using R packages

```{r}
#library()
library(tidyverse)

help("select")
```


## Importing Data in R

```{r}

# comma separated values, .csv
# from the base utils functions
employee_data_csv <- read.csv(file = "data/employee_data.cvs")

# from readr package within tidyverse package
employee_data_csv <- read_csv(file = "data/employee_data.cvs")

# from readr package within tidyverse package
employee_data <- read_rds('data/employee_data.rds')

# View data
employee_data
```


### Subsetting data

```{r}
# return first 10 rows with all columns
employee_data[1:10, ]

#employee_data[rows, columns]
employee_data[1:4, c(1:3)]

```


## Manipulating data

### using dplyr 
  1. Selecting variables(column)
```{r}
select(employee_data, 1:5)

head(employee_data, 10)
```

```{r}
select(employee_data, 1,3,5)
```


```{r}
select(employee_data, left_company, job_level)
```

Excluding variables

```{r}
select(employee_data, -left_company, -job_level)
```
  
  
```{r}
select(employee_data, -c(1:10))
```
  
```{r}

```
  
  2. Filtering data
```{r}
# employees that left the company
filter(employee_data, left_company == 'Yes') 
```


```{r}
# View employees that left from the Sales department
filter(employee_data, left_company == 'Yes', department == 'Sales') 
```


```{r}
# employees from Sales or Marketing department
filter(employee_data, department == 'Sales' | department == 'Marketing')
```

  3. Mutate
  
```{r}
# Create salary_scaled variable
# assuming salary is annual
mutate(employee_data, weekly_salary = salary/52)

mutate(employee_data, weekly_salary = salary/52)
```
  
  4. Summarize data
```{r}
# Let's get the average salary
summarise(employee_data, average_salary = mean(salary))
```
  

## Visualization data
### ggplot2 packages

```{r}
library(tidyverse)

heart_df <- read_rds('data/heart_disease.rds')
```


```{r}
# ggplot(data = [DATA], mapping = aes([MAPPING]) +
#          [GEOM_FUNCTION]() +
#          [FACET_FUNCTION]() +
#          [COORDINATE_FUNCTION]() +
#          labs(title = [TITLE],
#               x = [X AXIS LABEL],
#               y = [Y AXIS LABEL])
```


### Scatter plot

```{r}
ggplot2::ggplot(data = heart_df,mapping = aes(x = age,y = cholesterol))+
  geom_point()

?ggplot
```


```{r}
ggplot(data = heart_df, 
       mapping = aes(x = age, y = cholesterol)) +
       geom_point()
```

  - adding more variables
  
```{r}
ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
       geom_point()
```

  - faceting
  
```{r}
ggplot(data = heart_df, mapping = aes(x = age, y = cholesterol, color = heart_disease)) +
      geom_point() +
      facet_wrap(~ heart_disease, nrow = 1)
```


### Bar chart
```{r}
ggplot(data = heart_df, mapping = aes(x = heart_disease, fill = heart_disease)) +
       geom_bar(stat = "count") + 
       labs(title = "Heart Disease Prevalence", x = "Heart Disease Status",
            y = "Number of Patients")
```

### Histogram
```{r}
ggplot(data = heart_df, mapping = aes(x = resting_blood_pressure, fill = heart_disease)) +
       geom_histogram(color = "white", bins = 15) + 
       facet_wrap( ~ heart_disease, nrow = 1) +
       labs(title = "Distribution of Resting Blood Pressure",
            x = "Resting Blood Pressure",
            y = "Number of Patients")
```


## Statistical Analysis

  - Descriptive
  - Predictive
  - Prescriptive
  
## Data Science

### Linear regression - supervised learning

```{r}

diamonds_df <- diamonds

diamonds.lm <- lm()
diamonds.lm <- lm(formula = price ~ carat + cut + depth + clarity + color,
                  data = diamonds_df)

# Print summary statistics from diamond model
summary(diamonds.lm)

# Which components are in the regression object?
names(diamonds.lm)

# The coefficients in the diamond model
diamonds.lm$coefficients

# Coefficient statistics in the diamond model
summary(diamonds.lm)$coefficients

# Plot the relationship between true diamond values
#   and linear model fitted values

plot(x = diamonds$price,                          # True values on x-axis
     y = diamonds.lm$fitted.values,               # fitted values on y-axis
     xlab = "True Values",
     ylab = "Model Fitted Values",
     main = "Regression fits of diamond values")

abline(b = 1, a = 0)                             # Values should fall around this line!

# Create a dataframe of new diamond data
# carat + cut + depth + clarity + color
diamonds.new <- data.frame(carat = c(0.23, 0.36, 0.24),
                           cut = c("Very Good", "Very Good", "Ideal"),
                           depth = c(61.9, 55.7, 63.2),
                           clarity = c("SI2", "VS1", "I1"),
                           color = c("J", "E", "F"))

# Predict the value of the new diamonds using
#  the diamonds.lm regression model
predict(object = diamonds.lm,     # The regression model
        newdata = diamonds.new)   # dataframe of new data


```



### Thank you


